Survival Compass

### 2024-04-15 paper/paper.qmd
cleaning up data section of paper

---
title: "Survival Compass"
subtitle: "Statistical Insights into Lung Cancer Patients Journey Post Diagnosis"
author: 
  - Lexi Knight
thanks: "Code and data are available at: https://github.com/LexiKnight/Lung_Cancer/tree/main"
date: today
date-format: long 
abstract: "This study investigates the impact of pathogenic stage and treatment modalities on lung cancer survival post-diagnosis. Analysis of patient data reveals significant correlation between pathogenic stage, seeking treatment and survival outcomes. Notably, patients at advanced stages with metastases in distant sites beyond the lung, extensive lymph node involvement and tumors with extensive growth, invading nearby structures demonstrate lower survival rates. These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Our data is

```{r}
#| label: fig-bills
#| fig-cap: bb
#| echo: false

@@

@@

# Create bar chart for treatment type
treatment_type_counts <- analysis_data %>%
  group_by(treatment_type) %>%
  summarise(count = n ()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  arrange(desc(treatment_type))

ggplot(treatment_type_counts, aes(x = reorder(treatment_type, -percentage), y = percentage)) +
  geom_bar(stat = "identity", fill = "orange") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.2, color = "black", size = 3) +
  geom_text(data = treatment_type_counts, aes(label = treatment_type), hjust = 1.2, color = "black", size = 3) +
  labs(title = "Percentage of Patients by Treatment Type",
       x = "Treatment Type",
       y = "Percentage") +
  theme_minimal() +
  coord_flip() +
  scale_x_discrete(breaks = seq(0, max(treatment_type_counts$percentage), by = 5)) +
  theme(
    text = element_text(family = "patrick") # Set font to Patrick Hand
  )

@@

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

@@

We employ a linear regression model to predict the number of days from diagnosis to death for each lung cancer patient. The model is defined as follows:

\[
y_i \mid \mu_i, \sigma \sim \text{Normal}(\mu_i, \sigma)
\]

where:
\begin{itemize}
  \item $y_i$ represents the number of days from diagnosis to death for patient $i$.
  \item $mu_i$ denotes the expected number of days to death for patient $i$.
  \item $σ$ represents the standard deviation of the survival times.
\end{itemize}

The linear predictor $mu_i$ is specified as:

\begin{align*}
y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_{\text{pathologic\_stage}} \times \text{pathologic\_stage}_i \\
&\quad + \beta_{\text{lymph\_node}} \times \text{lymph\_node\_involvement}_i \\
&\quad + \beta_{\text{metastasis}} \times \text{presence\_of\_distant\_metastasis}_i \\
&\quad + \beta_{\text{tumor\_size}} \times \text{tumor\_size}_i \\
&\quad + \beta_{\text{treatment\_type}} \times \text{treatment\_type}_i
\end{align*}

where: 
\begin{itemize}
  \item \alpha represents the intercept term, capturing the baseline number of days to death.
  \item \beta_{\text{pathologic_stage}}, \beta_{\text{lymph_node}}, \beta_{\text{metastasis}}, \beta_{\text{tumor_size}}, \beta_{\text{treatment_type}} are the coefficients associated with each predictor variable.
\end{itemize}

@@

We run the model in R [@citeR] estimating the model coefficients ($/alpha$ and $/beta$) using Bayesian inference via the 'stan_glm()' function from the @rstanarm package. This approach leverages Markov Chain Monte Carlo (MCMC) algorithms to obtain posterior distributions for the model parameters, enabling robust estimation of parameter uncertainties and inference on the effects of predictor variables.

>>>>

---
title: "Survival Compass"
subtitle: "Statistical Insights into Lung Cancer Patients Journey Post Diagnosis"
author: 
  - Lexi Knight
thanks: "Code and data are available at: https://github.com/LexiKnight/Lung_Cancer/tree/main"
date: today
date-format: long 
abstract: "This study investigates the impact of pathogenic stage and treatment modalities on lung cancer survival post-diagnosis. Analysis of patient data reveals significant correlation between pathogenic stage, seeking treatment and survival outcomes. Notably, patients at advanced stages with metastases in distant sites beyond the lung, extensive lymph node involvement and tumors with extensive growth, invading nearby structures demonstrate lower survival rates. These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

## Software and R-packages
This project was created using statistical software, `R` [@citeR]. For data cleaning and manipulation, we used `tidyverse` [@tidy] package. For data visualization such as creating the figures, we used `ggplot2` [@ggplot2]. For converting into Parquet file, we used `arrow` [@arrow], managing file paths `here` [@here], statistical modeling `rstanarm` [@rstanarm]. Furthermore, `showtext` [@showtext] was used for aesthetic purposes.

## Methodology

The data for this study were collected from a comprehensive database comprising 981 lung cancer patients diagnosed between 1991 to 2013 from `Center for Open Science` [@cfos], a dataset acquired in Sydney Australia. The dataset included information on patient demographics, clinical characteristics, treatment modalities, and survival outcomes.

### Data Collection
We obtained data on lung cancer patients meeting the following criteria: histologically confirmed lung cancer diagnosis, availability of complete clinical data, treatment-naive patients, single malignancy and located in Australia. Patients with missing or incomplete information were excluded from the analysis.

### Data Cleaning
After obtaining the dataset, there were a series of rigorous preprocessing steps undertaken to ensure data quality and consistency. This included selecting the columns of interest namely; days to death post diagnosis, presence of distant metastasis, lymph node involvement, pathogenic stage, tumor size and treatment type. Next, we cleaned the column names and excluded missing values. Additionally, we converted the days to death column to numeric.Tests where included to ensure accuracy, reliability and validity of the datset for subsequent analysis and interpretation.

### Data Analysis
Descriptive and inferential statistical analyses were conducted to explore the dataset and derive meaningful insights. These included linear regression modeling.

## Features
The dataset comprised several key features relevant to lung cancer prognosis, including pathogenic stage, presence of distant metastasis, lymph node involvement, tumor size, and treatment type.

### Days to Death

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-one
#| fig-cap: Distribution of survival time in patients

@@

One of the critical features analyzed in this study is the duration between the date of lung cancer diagnosis and the date of death, referred to as "days to death." This metric serves as a key indicator of patient survival and provides valuable insights into the disease trajectory and prognosis. By examining the distribution of survival times among lung cancer patients post-diagnosis, we aim to characterize the temporal patterns of disease progression and assess the impact of various clinical and demographic factors on survival outcomes. Understanding the time course from diagnosis to death is crucial for guiding treatment decisions, predicting patient outcomes, and identifying opportunities for intervention to improve survival rates. Through comprehensive analysis of days to death data, we seek to elucidate the factors influencing patient survival in lung cancer and contribute to the refinement of prognostic models for clinical practice.

### Pathogenic Stage
Pathogenic stage, a critical determinant of lung cancer prognosis, was classified according to the [mention staging system used, e.g., TNM classification]. As depicted in Figure 1, the distribution of patients varied across different stages, with [mention any notable trends, e.g., a higher proportion of patients diagnosed at advanced stages].

### Presence of Distant Metastasis
The presence of distant metastasis, indicative of disease spread beyond the primary tumor site, significantly influences treatment decisions and patient outcomes. Figure 2 illustrates the percentage of patients with and without distant metastasis, highlighting the impact of metastatic disease on prognosis.

### Lymph Node Involvement
Lymph node involvement is a key prognostic factor in lung cancer, reflecting the extent of disease spread to regional lymph nodes. Figure 3 presents the distribution of patients based on lymph node involvement, demonstrating its association with survival outcomes.

### Tumor Size
Tumor size, often measured as the diameter of the primary tumor, is closely linked to disease progression and treatment response in lung cancer patients. Figure 4 showcases the distribution of patients across different tumor size categories, elucidating its significance as a prognostic factor.

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-bills
#| fig-cap: four

@@

@@

@@

We employ a linear regression model to predict the number of days from diagnosis to death for each lung cancer patient. The model is defined as follows:
\begin{align*}
y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma)
\end{align*}
where:

- $y_i$ represents the number of days from diagnosis to death for patient $i$.
- $\mu_i$ denotes the expected number of days to death for patient $i$.
- $\sigma$ represents the standard deviation of the survival times.

The linear predictor $\mu_i$ is specified as:
\begin{align*}
y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_{\text{pathologic\_stage}} \times \text{pathologic\_stage}_i \\
&\quad + \beta_{\text{lymph\_node}} \times \text{lymph\_node\_involvement}_i \\
&\quad + \beta_{\text{metastasis}} \times \text{presence\_of\_distant\_metastasis}_i \\
&\quad + \beta_{\text{tumor\_size}} \times \text{tumor\_size}_i \\
&\quad + \beta_{\text{treatment\_type}} \times \text{treatment\_type}_i
\end{align*}
where:

- $\alpha$ represents the intercept term, capturing the baseline number of days to death.
- $\beta$ {\text{pathologic\_stage}}, $\beta$ {\text{lymph\_node}}, $\beta$ {\text{metastasis}}, $\beta$ {\text{tumor\_size}}, $\beta$ {\text{treatment\_type}} are the coefficients associated with each predictor variable.

@@

We run the model in R [@citeR] estimating the model coefficients ($\alpha$ and $\beta$) using Bayesian inference via the 'stan_glm()' function from the @rstanarm package. This approach leverages Markov Chain Monte Carlo (MCMC) algorithms to obtain posterior distributions for the model parameters, enabling robust estimation of parameter uncertainties and inference on the effects of predictor variables.
